Theories of deliberated choice

### 2023-06-16 simple.tex
Falsifiability again

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N^*: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path. Its length is denoted by $\card{\hist} \in \N$ and its elements, $\hist(k)$ or $\hist_k$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist(k) = \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

@@

Say that $\gamma$ has depth at least $k \in \N^*$ iff $\exists (\alpha_i, 1 ≤ i ≤ k) \suchthat \forall 1 ≤ i ≤ k, \dg(\set{\alpha_j \suchthat 1 ≤ j ≤ i - 1}) \subsetneq \dg(\set{\alpha_j \suchthat 1 ≤ j ≤ i}) \land \forall 1 ≤ i ≤ k - 1, \alpha_i \cup \set{\dg(\alpha_i)} \subseteq \alpha_{i + 1}$.
The bounded depth condition on the defense relation imposes that for some $k \in \N^*$, $\gamma$ does not have depth at least $k$.
The depth of a theory is defined as the smallest value $k \in \N$ such that it does not have depth at least $k + 1$.
For example, $\gamma$ has depth $0$ iff $\dg = \emptyset$; and $\gamma$ has depth $1$ iff $[\exists \alpha_1 \suchthat \dg(\alpha_1) ≠ \emptyset] \land [\forall \alpha_1, \alpha_2 \suchthat \dg(\alpha_1) ≠ \emptyset \land \alpha_1 \cup \set{\dg(\alpha_1)} \subseteq \alpha_2: \dg(\alpha_2) \subseteq \dg(\alpha_1)]$.

@@

@@

	Imagine $i$ thinks as follows. $a_1$ counter $a_5$; $a_2$ counters $a_4$; but $a_4$ and $a_5$ together make $a_1$ and $a_2$ invalid, so that $(a_4, a_5)$ counter both $a_1$ and $a_2$ (and also $a_1$ and $a_2$ together). Also assume that $i$ has short memory so that $a_1, a_4, a_5$ does not convince ($a_4$ and $a_5$ win) while $a_4, a_5, a_1$ do ($a_1$ wins as $a_4$ is too far away). BUT $a_5, a_4, a_1$ should then not convince.
}

>>>>

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N^*: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path. Its length is denoted by $\card{\hist} \in \N$ and its elements, $\hist(k)$ or $\hist_k$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist(k) = \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$. A path $\hist$ is a sub-path of another path $\histp$ whenever $\hist \subseteq \histp$, thus, $\forall k \in \intvl{1, \card{\hist}}: \hist_k = \histp_k$.

@@

\subsection{Form of theories (draft)}
Given an exposure protocol $(I, \allargs, {\leadsto})$, we want to phrase some claims under the form of a logical sentence $C$ such that:
if $C$ holds, then it implies something about the dj of some $i$;
if $C$ fails, it is possible to falsify it.

We can observe set $O$ that forms a function $f(i) = \set{\alpha \ileadsto \phi, \alpha' \ileadsto ¬\phi, …}$, with $\alpha \subset \alpha \subset …$ and with finite $O$.
Our second condition is that $¬C \implies \exists O \suchthat O \implies ¬C$.
For example, if $C$ is $\alpha \ileadsto \phi \lor \beta \leadsto[j] \phi$, $O$ is $\alpha \ileadsto ¬\phi \land \beta \leadsto ¬\phi$, then $O$ implies $¬C$.
As another example, $C$ being $\exists \alpha \suchthat \alpha \ileadsto \phi$ is unfalsifiable, as its negation is $\forall \alpha: \alpha \nileadsto \phi$, which requires infinite $O$.

Consider $\leadsto$ such that $\exists \ar, \phi \suchthat \forall i: \ar \ileadsto \phi$.
Then some suitable $C$ exists.
It is $\forall i: \ar \ileadstost \phi$.
If false, it is falsifiable.
Even though it is not provable by any $O$.

Now consider $\leadsto$ such that $\exists J, K \subset I \suchthat \ar \ileadstost[j] \phi, \ar' \ileadstost[k] \phi$.
Then no suitable $C$ exists?

If $C$ has the form, whenever $a$ fails, $a'$ is then decisive, it is falsifiable, but does not lead to a decisive argument.
Want to show: $j \ileadstost \phi$. Need to show: $ar$ or $ar'$ is decisive.
Assume we know who is a $j$. Then we can!
Claim: $\ar \ileadsto \phi$.
Now assume that our claim has to be anonymous.
Also, assume the formulas involve one $i$ at a time (not needed, probably, but not much restrictive so okay for now).
So it is of the form $(i, \alpha, \phi) \lor (i, \alpha, \phi') \land (i, \alpha', \phi)…$ implies something decisive.
So it must be of the form: all $\alpha$ ending with $\ar$ do $\ileadsto \phi$. Or must imply this.
Let’s make it very strong, as much as possible, thus equal to $\ileadsto \cap \ileadsto[j]$.
That’s not strong enough!

Escaping?
Assume that we know that 50\% are J and 50\% are K.

\subsection{Form of theories}
Given an exposure protocol $(I, \allargs, {\leadsto})$, recall that ${\leadsto} \in P^{I × \allhist}$.

Given a pair $(I, \allargs)$, a claim $C$ is a set of possibilities for $\leadsto$, thus, $C \subseteq P^{I × \allhist}$.
The claim holds iff ${\leadsto} \in C$.
When writing $C$ as a formula, it is intended to mean the set $\set{{\leadsto} \in P^{I × \allhist} \suchthat C({\leadsto})}$.

Given a pair $(I, \allargs)$, a general observation $O$ is a finite subset of $I × \allhist × P$ such that $\forall i \in I, (i, \hist, p), (i, \histp, q) \in O: \card{\hist} ≤ \card{\hist'} \implies \hist \subseteq \hist'$.
General observations are the a priori possible observations, that is, those that can be considered possible when nothing is known about $\leadsto$.
Let $S_O = \set{f \in P^{I × \allhist} \suchthat O \subseteq f}$ denote the possibilities for $\leadsto$ compatible with the observations.

A claim $C$ is falsifiable iff $\exists O \suchthat O \implies ¬C$, by which I mean $S_O \cap C = \emptyset$.
\begin{example}
	$(\ar) \ileadsto \phi \lor (\ar, \ar) \ileadsto \phi \lor …$ is not falsifiable.
%	Its negation is $\forall k: (\ar^k) \nileadsto \phi$.
%	The claim represents the set $\set{{\leadsto} \in P^{I × \allhist} \suchthat (\ar) \ileadsto \phi \lor  (\ar, \ar) \ileadsto \phi \lor …}$.
	Given any $O$, define $l = \max_{\hist \suchthat \exists p \in P \suchthat (i, \hist, p) \in O} \card{\hist} + 1$. Pick any $\leadsto$ such that $O \subset {\leadsto} \land (\ar)^l \ileadsto \phi$. We have ${\leadsto} \in S_O \cap C ≠ \emptyset$.
\end{example}
\begin{example}
	$(\ar) \ileadsto \phi \lor (\ar') \ileadsto \phi$ is not falsifiable.
	Given any $O$, it may not contain both $((\ar), i, \phi)$ and $((\ar'), i, \phi)$, thus letting $t$ denote one of these triples not in $O$, pick any $\leadsto$ such that $O \subset {\leadsto} \land t \in {\leadsto}$. We have ${\leadsto} \in S_O \cap C ≠ \emptyset$.
\end{example}
\begin{example}[Fals. but shouldn’t]
	$\hist \ileadsto[j] \phi \land [(\ar) \ileadsto \phi \lor (\ar, \ar) \ileadsto \phi \lor …]$ is falsifiable.
	Consider $O = \set{(j, \hist, ¬\phi)}$.
\end{example}
\begin{example}[Fals. but shouldn’t, better]
	\label{ex:fbs}
	$\hist \ileadsto[j] \phi \land [(\ar) \ileadsto \phi \lor (\ar') \ileadsto \phi]$ is falsifiable.
	Consider $O = \set{(j, \hist, ¬\phi)}$.
\end{example}
\begin{example}
	$\forall i: \ar \ileadstost \phi$ is falsifiable.
	Consider any $i \in I$ and $O = \set{(i, (\ar), ¬\phi)}$.
\end{example}

\Cref{ex:fbs} shows that this definition of falsifiability is wrong.
We now say that $C$ is falsifiable iff $\forall {\leadsto} \in P^{I × \allhist}: ¬C({\leadsto}) \implies \exists O \subseteq {\leadsto} \suchthat O \implies ¬C$, meaning that
$\forall {\leadsto} \notin C: \exists O \subseteq {\leadsto} \suchthat S_O \cap C = \emptyset$.
\begin{example}
	$(\ar) \ileadsto \phi \lor  (\ar, \ar) \ileadsto \phi \lor …$ is falsifiable.
	Consider any ${\leadsto} \notin C$, thus satisfying $\exists k \suchthat (\ar)^k \nileadsto \phi$. Consider $O = \set{(i, (\ar)^k, {\leadsto}(i, (\ar)^k) ≠ \phi)}$, thus with $O \subset {\leadsto}$. We have $S_O \cap C = \emptyset$: consider any $O \subset f$, it is not in $C$.
\end{example}
\begin{example}
	$(\ar) \ileadsto \phi \lor (\ar') \ileadsto \phi$ is not falsifiable.
	Consider any ${\leadsto} \notin C$, thus satisfying $(\ar) \nileadsto \phi \land (\ar') \nileadsto \phi$.
	Consider any $O \subseteq {\leadsto}$ and let us prove that $S_O \cap C ≠ \emptyset$.
	By definition of a general observation, $O$ may not contain both $((\ar), i, \phi)$ and $((\ar'), i, \phi)$, thus letting $t$ denote one of these triples not in $O$, pick any $f$ such that $O \subset {f} \land t \in {f}$. We have ${f} \in S_O \cap C ≠ \emptyset$.
\end{example}
\begin{example}[TODO]
	$\hist \ileadsto[j] \phi \land [(\ar) \ileadsto \phi \lor (\ar, \ar) \ileadsto \phi \lor …]$ is not falsifiable?
	Consider any ${\leadsto} \notin C$, thus satisfying $\hist \nileadsto[j] \phi \lor [(\ar) \nileadsto \phi \land (\ar, \ar) \nileadsto \phi \land …]$.
	Consider $O = \set{(j, \hist, ¬\phi)}$.
\end{example}

@@

Say that $\gamma$ has depth at least $k \in \N^*$ iff $\exists (\alpha_i, 1 ≤ i ≤ k) \suchthat \forall 1 ≤ i ≤ k, \dg(\set{\alpha_j \suchthat 1 ≤ j ≤ i - 1}) \subsetneq \dg(\set{\alpha_j \suchthat 1 ≤ j ≤ i}) \land \forall 1 ≤ i ≤ k - 1, \alpha_i \cup \dg(\alpha_i) \subseteq \alpha_{i + 1}$.
The bounded depth condition on the defense relation imposes that for some $k \in \N^*$, $\gamma$ does not have depth at least $k$.
The depth of a theory is defined as the smallest value $k \in \N$ such that it does not have depth at least $k + 1$.
For example, $\gamma$ has depth $0$ iff $\dg = \emptyset$; and $\gamma$ has depth $1$ iff $[\exists \alpha_1 \suchthat \dg(\alpha_1) ≠ \emptyset] \land [\forall \alpha_1, \alpha_2 \suchthat \dg(\alpha_1) ≠ \emptyset \land \alpha_1 \cup \dg(\alpha_1) \subseteq \alpha_2: \dg(\alpha_2) \subseteq \dg(\alpha_1)]$.

\commentOC{Intuition. Invoke $\dg$ at $\alpha$ only when $\alpha$ fails. Permit to invoke a previous defense again as maybe the counter was too long to perform when summarized. If $a_2$ fails, we must invoke $\dg(…, a_2, x)$; failures of $\dg(…, a_2, …, x)$ should not count. We should invoke $\dg$ only when $\alpha$ fails with some defense produced by $\dg()$ belongs to $\histend$.}

%Falsification seems done when $\exists \alpha, \beta \suchthat d(\beta) \in \histend \land \alpha \ileadsto p$. But then seems to be required to tell $d$ what has been tried already!}

@@

\commentOC{Condition: $\hist \nileadsto \pga \land \exists \hist' \subseteq \hist \suchthat \dg(\histp) \in \histend \land \dg(\hist)$ is not decisive ($\exists \beta \subseteq \hist \suchthat \dg(\hist) \in \beta_\text{end} \land \beta \ileadsto \pga$).}

@@

	Imagine $i$ thinks as follows. $a_1$ counter $a_5$; $a_2$ counters $a_4$; but $a_4$ and $a_5$ together make $a_1$ and $a_2$ invalid, so that $(a_4, a_5)$ counter both $a_1$ and $a_2$ (and also $a_1$ and $a_2$ together). Also assume that $i$ has short memory so that $a_1, a_4, a_5$ does not convince ($a_4$ and $a_5$ win) while $a_4, a_5, a_1$ do ($a_1$ wins as $a_4$ is too far away). BUT $a_5, a_4, a_1$ should then not convince.

The solution is: after $a1, a5, a2$, $a1$ appears decisive, thus also $a1,a5, a2, a1, a4$ ok. Then some subset $a$ of $a1 \cup a_5 \cup a2$ should satisfy $a, a4$ OK and replace $a1$. And $a$ is decisive.
}
